dispRity manual

### 2017-11-10 paper/response_to_reviewers.tex

\item{\textcolor{blue}{Line 4 – changes ``in'' to ``to''. Line 6 – changes ``axis'' to ``axes''}}
\label{abstract_typo}

I have fixed these typos.

\item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}

In \texttt{geomorph}, the procrustes variance is defined as the sum of the diagonal covariance matrix between each elements divided by the number of elements (or effectively the sum of the squared ordination values divided by the number of elements - see example in \texttt{R} for \texttt{?dispRity::geomorph.ordination}).
Using this disparity metric is appropriate for testing volume based hypothesis (e.g. Does group A occupies more morphospace than group B?) but does not allow to test position based hypothesis (e.g. Does group A occupies a different part of the morphospace than group B).
For example, one might find out that group A and B have the same volume but that doesn't tell us if they occupy the same position in the morphospace.

I've highlighted this in the text:

\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests(Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017). This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).}

### 2017-11-10 inst/gitbook/05_guts.Rmd
Updated manual character simulation
needs to do morphospace simulations

>>>>

## The `dispRity` object content
The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:

 * `$matrix`: an object of class `matrix`, the full multidimensional space.
 * `$call`: an object of class `list` containing information on the `dispRity` object content.
 * `$subsamples`: an object of class `list` containing the subsamples of the multidimensional space.
 * `$disparity`: an object of class `list` containing the disparity values.

The `dispRity` object is loosely based on `C` structure objects.
In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subsample.

### `$matrix`
This is the multidimensional space, stored in the `R` environment as a `matrix` object.
It requires row names but not column names.
By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
This element of the `dispRity` object is never modified.

### `$call`
This element contains the information on the `dispRity` object content.
It is a `list` that can contain the following:

 * `$call$subsamples`: a vector of `character` with information on the subsamples type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"punctuated"`, `"gradual"`). This element generated only once via `time.subsamples()` and `custom.subsamples()`.
 * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use. This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
 * `$call$bootstrap`: this is a `list` containing three elements:
    * `[[1]]`: the number of bootstrap replicates (`numeric`)
    * `[[2]]`: the bootstrap method (`character`)
    * `[[3]]`: the rarefaction levels (`numeric` vector)
 * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`. These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).

### `$subsamples`
This element contain the eventual subsamples of the multidimensional space.
It is a `list` of subsample names.
Each subsample name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
This `elements` matrix is the raw (observed) elements in the subsamples.
The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subsample).
Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsamples called `sub1`:

```
 disparity$subsamples$sub1$elements
      [,1]
 [1,]    5
 [2,]    4
 [3,]    6
 [4,]    7
```

The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
The following elements in `diparity$subsamples$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
The columns will correspond to different bootstrap replicates.
For example:

```
 disparity$subsamples$sub1[[2]]
      [,1] [,2] [,3] [,4]
 [1,]   57   43   70    4
 [2,]   43   44    4    4
 [3,]   42   84   44    1
 [4,]   84    7    2   10
```
This signifies that we have four bootstrap pseudo-replicates pointing each time to four elements in `$matrix`.
The next element (`[[3]]`) will be the same for the eventual first rarefaction level (i.e. the resulting bootstrap matrix will have _m_ rows where _m_ is the number of elements for this rarefaction level).
The next element after that (`[[4]]`) will be the same for with an other rarefaction level and so forth...

### `$disparity`
The `$disparity` element is identical to the `$subsamples` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:

```
 disparity$disparity$sub1$elements
      [,1]
 [1,]    1.82
```

This is the observed disparity (1.82) for the subsample called `sub1`.
If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:

```
 disparity$disparity$sub1$elements
      [,1]
 [1,]    0.82
 [2,]    2.82
```

The following elements in the list follow the same logic as before: rows are disparity values (one row for a dimension level 1 metric, multiple for a dimensions level 2 metric) and columns are the bootstrap replicates (the bootstrap with all elements followed by the eventual rarefaction levels).
For example for the bootstrap without rarefaction (second element of the list):

```
 disparity$disparity$sub1[[2]]
         [,1]     [,2]     [,3]     [,4]
[1,] 1.744668 1.777418 1.781624 1.739679 
```

### 2017-11-10 inst/gitbook/06_palaeo-demo.Rmd
Updated manual character simulation
needs to do morphospace simulations

This demo aims to give quick overview of the `dispRity` package (v.0.3) for palaeobiology analyses of disparity, including disparity through time analyses.
Please refer to my [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html) that explains the functions in more detail.

In brief, this package allows users to easily perform **disparity-through-time** analysis.
This type of analysis often starts by ordinating **morphometric** or **cladistic** data into a multidimensional object hereafter called the **morphospace**.
One might be interested in studying how the occupancy of the morphospace has changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
Disparity can then be seen as a value or a distribution of values summarising the diversity of morphologies.

This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years.

@@

### Installing `dispRity`

You can install this package easily if you are using the latest version of `R` and `devtools`.

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
```

@@

In this example, we are going to use a subset of the data from [@beckancient2014]. 
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).

@@

```{r, eval = FALSE}
if(!require(strap)) install.packages("strap")
library(strap)
geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
```

@@

The output object is a `dispRity` object.
For details about this object class, please refer to the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
In brief, however, `dispRity` objects are lists of different elements (i.e. disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.

@@

```{r}
## Bootstrapping each time subsample 100 times
(boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100))

## Getting the minimum number of rows (i.e. taxa) in the time subsamples
min(unlist(lapply(boot_bin_morphospace$subsamples, lapply, nrow)))

@@

We can now calculate the disparity within each time subsamples along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
For more details, please refer to the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).

In this example, we are going to calculate the spread of the data in each time subsample by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subsample using the `dispRity` function.
Thus, in this example, disparity is defined by the multi-dimensional variance of each time subsample (i.e. the spread of the taxa within the morphospace).
Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
We use this here because it is a standard metric used in disparity-through-time analysis.
<!-- NC: reference? -->

@@

### References 
<!-- Add these -->

>>>>

This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.

This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).

@@

@@

In this example, we are going to use a subset of the data from Beck and Lee (2014).
See the [example data](#example-data) description for more details.
Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).

@@

```{r, eval = TRUE}
if(!require(strap)) install.packages("strap")
strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
```

@@

The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
In brief, however, `dispRity` objects are lists of different elements (i.e. disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.

@@

```{r}
## Bootstrapping each time subsample 100 times (default)
(boot_bin_morphospace <- boot.matrix(binned_morphospace))

## Getting the minimum number of rows (i.e. taxa) in the time subsamples
min(size.subsamples(boot_bin_morphospace))

@@

We can now calculate the disparity within each time subsamples along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
For more details, please refer to the [`dispRity` metric section](#disparity-metrics).

In this example, we are going to calculate the spread of the data in each time subsample by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subsample using the `dispRity` function.
Thus, in this example, disparity is defined by the multi-dimensional variance of each time subsample (i.e. the spread of the taxa within the morphospace).
Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
We use this here because it is a standard metric used in disparity-through-time analysis (Wills _et al._ 1994).

@@

### 2017-11-10 inst/gitbook/index.Rmd
Updated manual character simulation
needs to do morphospace simulations

## What is `dispRity`?
This is a package for measuring disparity in `R`.
It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
However, there are many more applications in ecology, evolution and beyond.

@@

You can install this package easily if you use the latest version of `R` (> 4.0) and `devtools`.

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dispRity)
## Setting the random seed for repeatability
set.seed(123)
```

>>>>

## What is `dispRity`?
This is a modular package for measuring disparity in `R`.
It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
However, there are many more applications in ecology, evolution and beyond.

### Modular?
Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
This will allow you to finely tune your multidimensional analysis to the needs of your specific question!

@@

You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dispRity)
## Setting a random seed for repeatability
set.seed(123)
```
